---
title: "Assignment 2"
author: 
  - Simon Alain Roger Gerdolle 23-611-940
  - Leonardo Gonnelli 23-617-111
  - Cédric Ly 23-615-487
date: "2025-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Import packages
```{r, ECHO = TRUE}
library(readr)
library(ggplot2)
library(dplyr)
```

# Load the data sets
```{r}
cleaned <- read_delim("ps2_cantons_cleaned.csv", delim = ";")
View(cleaned)
surnames <- read_delim("ps2_cantons_surnames.csv")
View(surnames)
```

# 1. Data Preparation
```{r}
# Rename
surnames <- surnames %>%
  mutate(canton = if_else(canton == "Sankt Gallen", "St. Gallen", canton))

# Merge data sets
merged_data <- cleaned %>%
  left_join(surnames %>% select(canton, share_mueller), by = "canton") %>%
  select(-immigration)
View(merged_data)

```

```{r}
# data checks
skimr::skim(merged_data)
print(merged_data$canton)

```

# 2. Estimation of a Univariate Model
$$
gdp = \beta_0 + \beta_1 \cdot share \_mueller + u
$$
```{r}
model1 <- lm(gdp ~ share_mueller, data = merged_data)
print(model1$coefficients)
```

$$
gdp = 8059.703 + 804465.380 \cdot share \_mueller + u
$$
Interpretation :

1. Beta_⁠0 = 8059.703 (Intercept):
This represents the expected GDP per canton (in million CHF) when the share of the population with the surname “Mueller” is equal to 0.
In other words, if a canton had no residents with this surname, the model predicts that its GDP would be approximately 8059.703 million CHF.

2.	Beta_1 = 804465.380 (Slope):
This coefficient indicates that for every 1 percentage point increase in the share of the population with the surname “Mueller”, the model predicts that GDP will increase by 804,465.38 million CHF.
This is an extremely large effect, suggesting a very strong positive correlation between the share of people with the surname Mueller and GDP.

```{r}
#prediction of gpd with share_mueller = 0,1
B0 <-  8059.703
B1 <- 804465.380
X1 <- 0.1

y = B0 + B1 * X1
```

# 3. Comparison of two Univariate Models
```{r}
#regression of gdp on population
model2 <- lm(gdp ~ population, data = merged_data)
print(model2$coefficients)
```

$$
gdp = -1013.109 + 88621.379 \cdot population + u
$$

```{r}
# R^2 Share_mueller
gdp_pred <- model1$coefficients[1] + model1$coefficients[2] * merged_data$share_mueller
SSR1 <- sum((merged_data$gdp - gdp_pred)^2)
SST1 <- sum((mean(merged_data$gdp) - merged_data$gdp)^2)
SSE1 <- SST1 - SSR1
rsquared1 <- 1 - SSR1/SST1
print(rsquared1)

# R^2 from lm model summary object
print(summary(model1)$r.squared)


# R^2 Population
gdp_pred <- model2$coefficients[1] + model2$coefficients[2] * merged_data$population
SSR2 <- sum((merged_data$gdp - gdp_pred)^2)
SST2 <- sum((mean(merged_data$gdp) - merged_data$gdp)^2)
SSE2 <- SST2 - SSR2
rsquared2 <- 1 - SSR2/SST2
print(rsquared2)

# R^2 from lm model summary object
print(summary(model2)$r.squared)

```
*Interpretation*
	•	The regression using population as the independent variable explains 93.5% of the variation in GDP, which is significantly higher than the 66.8% explained by share_mueller.
	•	This suggests that population is a much stronger predictor of GDP than share_mueller.

  *Unbiasedness*
	•	The unbiasedness of an estimator depends on whether the regression assumptions hold (e.g., SLR.4: no omitted variable bias).
	•	Population is likely a fundamental determinant of GDP, meaning the coefficient in the second model is less prone to omitted variable bias.
	•	Share_mueller may correlate with other economic or demographic factors, leading to potential omitted variable bias in the first regression.

*Predictive Power and Unbiasedness*
	•	Higher R² means better predictive power, but it does not imply unbiasedness.
	•	If important variables are missing, even a high R² can be misleading due to omitted variable bias.
	•	Population seems to be a more natural predictor of GDP than share_mueller, making it a better choice for modeling.

# 4 Univariate model without a constant
$$
gdp = \beta_1 \cdot share \_mueller + u
$$
```{r}
#prediction beta_1
model3 <- lm(gdp ~ 0 + share_mueller, data = merged_data)
print(model3)
```
*Why Did the Coefficient Change?*
	•	Removing the intercept forces the regression line through (0,0), altering the slope.
	•	Without an intercept, \beta_1 is computed as \sum XY / \sum X^2, leading to a higher magnitude than in the model with an intercept.

*Is the Estimate Unbiased?*
	•	No, because GDP likely doesn’t drop to zero when share_mueller = 0.
	•	Omitting the intercept biases \beta_1 since it absorbs the missing constant term.
	•	A model with an intercept is usually more accurate.

# 5 Code your own OLS estimator
```{r}
# Sample code 
# Define your function 
my_ols_univariate <- function(y, x, add_constant = TRUE) {
  if (add_constant) {
    # Compute beta0 and beta1 
    
    beta1 = NA #### add your computation
    beta0 = NA #### add your computation
    
    return(list(beta0 = beta0, beta1 = beta1))
  } else {
    # Compute beta1
    
    beta1 = NA #### add your computation
    
    return(list(beta1 = beta1))
  }
}

# Call your function
coefficients <- my_ols_univariate(y=c(1,2,3),x=c(1,2,2),add_constant=TRUE)
print(coefficients)

```

```{r}
# Custom OLS function
ols_estimator <- function(y, x, add_constant = TRUE) {
  if (add_constant) {
    # Compute beta1 and beta0 (with intercept)
    beta1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
    beta0 <- mean(y) - beta1 * mean(x)
    return(list(beta0 = beta0, beta1 = beta1))
  } else {
    # Compute beta1 only (without intercept)
    beta1 <- sum(x * y) / sum(x^2)
    return(list(beta1 = beta1))
  }
}

# Extract variables
y <- merged_data$gdp
x <- merged_data$share_mueller

# Compute estimates
ols_with_const <- ols_estimator(y, x, add_constant = TRUE)
ols_no_const <- ols_estimator(y, x, add_constant = FALSE)

# Compare with lm() results
lm_with_const <- lm(gdp ~ share_mueller, data = merged_data)
lm_no_const <- lm(gdp ~ 0 + share_mueller, data = merged_data)

summary(lm_with_const)
summary(lm_no_const)
```
