---
title: "Assignment 5"
author:
  - Leonardo Gonnelli (23-617-111)
  - Cédric Ly (23-615-487)
  - Simon Gerdolle (23-611-940)
output: html_document
date: "2025-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(haven)
library(skimr)
library(AER)
```

Import the data:
```{r}
df <- read_csv("education_income-3.csv")
skimr::skim(df)
df_clean <- df %>%
  mutate(
    income = na_if(income, -9999999),
    hhincome = na_if(hhincome, -9999999),
    college = na_if(college, -1)
  ) %>%
  drop_na(income, hhincome, college, scholarship)
View(df)
```

# Part 0

As we are trying to understand the correlation (or maybe further the causality) between education and income, we will need to drop the observations where one of those variables are unknown. We see that there is 1340 observations where the income and college variables are unknown. **For now, we don't know if we need to drop the observations where hhincome is unknown**.

```{r}
df_clean <- df %>%
  mutate(
    income = na_if(income, -9999999),
    hhincome = na_if(hhincome, -9999999),
    college = na_if(college, -1)
  ) %>%
  drop_na(income, hhincome, college, scholarship)

skimr::skim(df_clean)
View(df_clean)
```
We can now proceed to the next task.

# Part 1.1

We want to investigate whether an increase in the scholarship value received by a student at the beginning at college has an effect on their income 10 years later. To do so, we estimate the following linear regression model where income 10 years later is regressed on the scholarship amount received.

$$income = \alpha_0 + \alpha_1 \cdot scholarship + u$$

```{r}
model <- lm(data = df_clean, income ~ scholarship)
summary(model)
```
Here, the intercept $\alpha_0$ shows the expected income for an individual who would have received 0USD of scholarship value, which is irrelevant for our analysis as we are working only on ex-students that received at least 10kUSD of scholarship value. To ensure the intercept is interpretable, we center the scholarship variable at its minimum observed value. 

```{r}
df_clean$scholarship_mincentered <- df_clean$scholarship - min(df_clean$scholarship)
```

```{r}
model <- lm(data = df_clean, income ~ scholarship_mincentered)

summary(model)
```
The results are the same, except we now have the exact intercept for the minimum observed scholarship value, i.e. 10,000 USD:

$\alpha_0$ = 125.351: the estimated income 10 years after high school graduation for a student who received the minimum scholarship amount (10,000 USD).

$\alpha_1$ = 0.14736: the estimated coefficient for scholarship indicates that for each additional 1,000 USD received, future income increases by approximately 161.45 USD.

As the p-value is below 0.001, the result is statistically significant. However, the R-squared is very low (R^2 = 0.0007), suggesting that scholarship alone explains only a small share of the variation in income—many other factors likely play a more important role.

# Part 1.2

For this task, we are asked to check whether the effect of a scholarship on income has changed over time. In order to do this, we need to estimate the following multivariate linear regression model :
$$ income = \alpha_0 + \alpha_1\cdot scholarship + \alpha_2 \cdot year + \alpha_3 \cdot (scholarship \times year) + u$$
**Null and alternative hypotheses**

- Null hypothesis $H_0$: $\alpha_3 = 0$, i.e. the effect of scholarship on income did not change over time.
- Alternative hypothesis $H_1$: $\alpha_3 \neq 0$, i.e. the effect of scholarship on income did change over time.

To estimate our linear regression model, we first need to center the year variable at 2000 (in order to get an interpretable intercept, otherwise our intercept $\alpha_0$ will give us the expected income at year 0 which is irrelevant for us):

```{r}
df_clean$year_mincentered <- df_clean$year - min(df_clean$year)
view(df_clean)
```

Then, we need to compute the interaction between the scholarship and the year variables:
```{r}
df_clean$interaction <- df_clean$scholarship * df_clean$year
view(df_clean)
```

Now, we an run our model:
```{r}
model_interaction <- lm(income ~ scholarship_mincentered + year_mincentered + interaction, data = df_clean)
summary(model_interaction)
```

We observe that the the estimated interaction $\alpha_3 = -0.001016$ with a se of 0.0093, a t-statistic of -0.109 and a **p-value of 0.9136**.
Because the p-value is very high compared to the conventional significance level (0.05), **we fail to reject the null hypothesis** $H_0$, there is no statistical evidence that the effect of scholarship on income has changed over time.

# Part 1.3
The goal of this analysis is to estimate the causal effect of obtaining a college degree on an individual's income ten years after high school graduation. The government seeks to understand whether higher education leads to improved economic outcomes, using data from a randomized scholarship program.

A naïve approach regresses income on college attendance using ordinary least squares (OLS). However, college is a choice variable, likely influenced by unobserved factors like motivation and ability, which also affect income. This introduces endogeneity, leading to omitted variable bias in the OLS estimate.

$$income = \beta_0 + \beta_1 \cdot college_i + \epsilon_i $$
However, college attendance is a choice variable likely influenced by unobserved factors such as motivation, ability, or family support, which also affect income. This raises endogeneity concerns, and any estimate of $\beta_1$ using OLS would be biased due to omitted variable bias.

To address this, we use a two-stage least squares (2SLS) approach, leveraging the randomized assignment of scholarship values as an instrumental variable (IV). Since scholarship amounts were randomly allocated among those who passed an initial eligibility screening, they serve as an exogenous source of variation in college attendance.

The two stages are specified as follows:

First Stage:
$$college_i = \pi_0 + \pi_1 \cdot scholarship_i + \upsilon_i $$
Second Stage:
$$income = \beta_0 + \beta_1 \cdot \hat{college}_i + \epsilon_i $$
$\hat{college}_i$ is the predicted probability of attending college based on the scholarship value. This approach identifies the Local Average Treatment Effect (LATE) — the causal effect of college for those individuals whose college attendance decisions were influenced by the amount of scholarship received.


```{r}
#OLS Estimate
ols_model <- lm(income ~ college, data = df_clean)
summary(ols_model)
```
Estimate: 41.51 (SE = 0.53, p < 2e-16)
*Interpretation:* 
College graduates earn $41,511 more annually than non-graduates, on average. However, this estimate may be upward biased due to unobserved factors like motivation or ability.

```{r}
#First Stage (Relevance Check)
first_stage <- lm(college ~ scholarship, data = df_clean)
summary(first_stage)
```
Estimate of $\pi_1$ (scholarship → college): 0.00958 (p < 2e-16)
F-statistic: 278.1
The instrument is strong and statistically significant, satisfying the relevance condition. The F-statistic far exceeds the conventional threshold of 10, indicating we are unlikely to face a weak instrument problem.

```{r}
iv_model <- ivreg(income ~ college | scholarship, data = df_clean)
summary(iv_model)
```
Estimate: 15.38 (SE = 4.49, p = 0.0006)
*Interpretation: *
Receiving a college degree causes a $15,381 increase in annual income for those influenced by the scholarship (the compilers). This estimate is considerably lower than the OLS estimate, suggesting that the OLS model may overstate the returns due to selection bias.

```{r}
iv_model_controls <- ivreg(income ~ college + hhincome + race | scholarship + hhincome + race, data = df_clean)
summary(iv_model_controls)
```
college coefficient: 21.00 (SE = 2.26, p < 2e-16)

These results strongly support the hypothesis that college education increases income. However, the comparison between OLS and IV estimates reveals important insights:

- OLS likely overstates the effect due to upward bias from omitted factors (e.g., ability, motivation).
- IV gives a more credible estimate of the causal effect by isolating variation in college attendance due to a randomized instrument.
- The LATE interpretation means this effect applies specifically to individuals whose decision to attend college was affected by the scholarship amount i.e., marginal students.

*Conclusion:*
Using a credible instrument — randomized scholarship values — we estimate that obtaining a college degree causally increases income by between 15,000USD and 21,000USD per year for marginal students. These results are statistically significant and robust to the inclusion of important controls.

This finding supports policies that lower financial barriers to higher education, especially for students on the margin of college attendance. However, policymakers should note that these benefits apply most clearly to individuals responsive to financial incentives and may not generalize to all students.