---
title: "Assignment 5"
author:
  - Leonardo Gonnelli (23-617-111)
  - Cédric Ly (23-615-487)
  -Simon Gerdolle (23-611-940)
output: html_document
date: "2025-05-09"
---

```{r}
library(tidyverse)
```

Import the data:
```{r}
df <- read_csv("education_income-3.csv")
skimr::skim(df)
```

*Part 0*

As we are trying to understand the correlation (or maybe further the causality) between education and income, we will need to drop the observations where one of those variables are unknown. We see that there is 1340 observations where the income is unknown. **For now, we don't know if we need to drop the observations where hhincome is unknown**.

```{r}
df_clean <- df %>%
  filter(income>=0)

skimr::skim(df_clean)
```
We can now proceed to the next task.

*Part 1*

We want to investigate whether an increase in the scholarship value received by a student at the beginning at college has an effect on their income 10 years later. To do so, we estimate the following linear regression model where income 10 years later is regressed on the scholarship amount received.

$$income = \alpha_0 + \alpha_1scholarship + u $$

```{r}
model <- lm(data = df_clean, income ~ scholarship)

summary(model)
```
Here, the intercept $\alpha_0$ shows the expected income for an individual who would have received 0USD of scholarship value, which is irrelevant for our analysis as we are working only on ex-students that received at least 10kUSD of scholarship value. To ensure the intercept is interpretable, we center the scholarship variable at its minimum observed value. 

```{r}
df_clean$scholarship_mincentered <- df_clean$scholarship - min(df_clean$scholarship)
```

```{r}
model <- lm(data = df_clean, income ~ scholarship_mincentered)

summary(model)
```
The results are the same, except we now have the exact intercept for the minimum observed scholarship value, i.e. 10,000 USD:

$\alpha_0$ = 125.148: the estimated income 10 years after high school graduation for a student who received the minimum scholarship amount (10,000 USD).

$\alpha_1$ = 0.16145: the estimated coefficient for scholarship indicates that for each additional 1,000 USD received, future income increases by approximately 161.45 USD.

As the p-value is below 0.001, the result is statistically significant. However, the R-squared is very low (R^2 = 0.0007), suggesting that scholarship alone explains only a small share of the variation in income—many other factors likely play a more important role.


 